Parameter Recovery Validation

### 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): generalize hyperparameter relationships based on T_M
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

- $0 < t^*_{on,i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
- $\delta^*_i < 2.0$ (short activation duration relative to process, assuming conservative prior Gamma(5,1))
- $R_{\text{on},i} > 2.0$ (sufficient fold-change to observe activation)

@@

- $0 < t^*_{on,i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
- $\delta^*_i > 2.5$ (long activation duration relative to process, assuming conservative prior Gamma(5,1))
- $R_{\text{on},i} > 2.0$ (strong fold-change during activation)

@@

With independent parameterization, gene timing parameters should be calibrated to $T_M$ expectations:

\begin{align}
t^*_{on,i} &\sim \text{Normal}(0.3 \times 5, \sigma_{on}^2) = \text{Normal}(1.5, \sigma_{on}^2) \quad \text{(expect activation at 30% of process)} \\
\delta^*_i &\sim \text{LogNormal}(\log(0.4 \times 5), \sigma_{\delta}^2) = \text{LogNormal}(\log(2.0), \sigma_{\delta}^2) \quad \text{(expect 40% duration activation)}
\end{align}

@@

# Prior hyperparameters (updated for independent absolute parameterization)
# Assuming T_M ~ Gamma(5.0, 1.0) with E[T_M] = 5.0
E_T_M = 5.0
priors = {
    'R_on': {'loc': 0.916, 'scale': 0.4},                           # log(2.5), fold-change (LogNormal)
    't_on_star': {'loc': 1.5, 'scale': 0.8},                       # Normal(1.5, 0.8²), absolute onset
    'delta_star': {'loc': np.log(2.0), 'scale': 0.45},             # log(2.0), absolute duration (LogNormal)
}

# Calculate constraint feasibility for each pattern (updated for independent parameterization)
patterns = {
    'pre_activation': [
        ('t_on_star', '<', 0.0),        # Activation before observation window
        ('R_on', '>', 2.0),             # Moderate to strong fold change
    ],
    'transient': [
        ('t_on_star', '>', 0.0),                    # Activation within observation window
        ('t_on_star', '<', 1.5),                   # Early enough to complete cycle
        ('delta_star', '<', 2.0),                  # Short pulse duration
        ('R_on', '>', 2.0),                        # Sufficient fold change to observe
    ],
    'sustained': [
        ('t_on_star', '>', 0.0),                    # Activation within observation window
        ('t_on_star', '<', 1.5),                   # Early activation onset
        ('delta_star', '>', 2.5),                  # Long activation duration
        ('R_on', '>', 2.0),                        # Strong fold change
    ]
}

@@

- **Pre-activation**: $t^*_{on,i} < 0.0$ and $R_{\text{on},i} > 2.0$
- **Transient**: $0 < t^*_{on,i} < 1.5$, $\delta^*_i < 2.0$, and $R_{\text{on},i} > 2.0$ (assuming conservative prior Gamma(5,1))
- **Sustained**: $0 < t^*_{on,i} < 1.5$, $\delta^*_i > 2.5$, and $R_{\text{on},i} > 2.0$ (assuming conservative prior Gamma(5,1))

@@

- **$t^*_{on,i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$**: Absolute activation onset time, typically centered around 1.5 to expect activation at 30% of process duration (assuming conservative prior Gamma(5,1) with mean 5). **Negative values** enable pre-activation scenarios where genes were already activated before observation began.

- **$\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$**: Absolute activation duration, typically centered around 2.0 to expect 40% duration activation (assuming conservative prior Gamma(5,1) with mean 5). This enables both transient (short duration) and sustained (long duration) activation patterns.

@@

- **Pre-activation** ($t^*_{on,i} < 0$): Probability depends on $\mu_{on}$ and $\sigma_{on}$ choices
- **Transient** ($0 < t^*_{on,i} < 1.5$ and $\delta^*_i < 2.0$): Balanced probability for observable complete cycles (assuming conservative prior Gamma(5,1))
- **Sustained** ($0 < t^*_{on,i} < 1.5$ and $\delta^*_i > 2.5$): Balanced probability for long-duration activation (assuming conservative prior Gamma(5,1))

>>>>

- $0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$ (early activation within process duration)
- $\delta^*_i < 0.4 \mathbb{E}[T_M]$ (short activation duration relative to process)
- $R_{\text{on},i} > 2.0$ (sufficient fold-change to observe activation)

@@

- $0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$ (early activation within process duration)
- $\delta^*_i > 0.5 \mathbb{E}[T_M]$ (long activation duration relative to process)
- $R_{\text{on},i} > 2.0$ (strong fold-change during activation)

@@

With independent parameterization, gene timing parameters should be calibrated to the expected process duration based on chosen $T_M$ hyperparameters:

**General principle:**
\begin{align}
t^*_{on,i} &\sim \text{Normal}(f_{on} \times \mathbb{E}[T_M], \sigma_{on}^2) \\
\delta^*_i &\sim \text{LogNormal}(\log(f_{\delta} \times \mathbb{E}[T_M]), \sigma_{\delta}^2)
\end{align}

where $f_{on} = 0.3$ (expect activation at 30% of process) and $f_{\delta} = 0.4$ (expect 40% duration activation) are scaling factors based on biological expectations.

**Concrete examples for different experimental contexts:**
- **Conservative (Gamma(5,1))**: $\mathbb{E}[T_M] = 5$ → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$
- **Skeptical (Gamma(1,1))**: $\mathbb{E}[T_M] = 1$ → $t^*_{on,i} \sim \text{Normal}(0.3, \sigma_{on}^2)$
- **Permissive (Gamma(1,0.2))**: $\mathbb{E}[T_M] = 5$ → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$

@@

# Prior hyperparameters (updated for independent absolute parameterization)
# Adaptive to chosen T_M hyperparameters
alpha_T, beta_T = 5.0, 1.0  # Conservative prior example
E_T_M = alpha_T / beta_T  # Expected process duration
priors = {
    'R_on': {'loc': 0.916, 'scale': 0.4},                           # log(2.5), fold-change (LogNormal)
    't_on_star': {'loc': 0.3 * E_T_M, 'scale': 0.8},               # Normal(0.3*E[T_M], 0.8²), absolute onset
    'delta_star': {'loc': np.log(0.4 * E_T_M), 'scale': 0.45},     # log(0.4*E[T_M]), absolute duration (LogNormal)
}

# Calculate constraint feasibility for each pattern (updated for independent parameterization)
patterns = {
    'pre_activation': [
        ('t_on_star', '<', 0.0),        # Activation before observation window
        ('R_on', '>', 2.0),             # Moderate to strong fold change
    ],
    'transient': [
        ('t_on_star', '>', 0.0),                    # Activation within observation window
        ('t_on_star', '<', 0.3 * E_T_M),           # Early enough to complete cycle
        ('delta_star', '<', 0.4 * E_T_M),          # Short pulse duration
        ('R_on', '>', 2.0),                        # Sufficient fold change to observe
    ],
    'sustained': [
        ('t_on_star', '>', 0.0),                    # Activation within observation window
        ('t_on_star', '<', 0.3 * E_T_M),           # Early activation onset
        ('delta_star', '>', 0.5 * E_T_M),          # Long activation duration
        ('R_on', '>', 2.0),                        # Strong fold change
    ]
}

@@

- **Pre-activation**: $t^*_{on,i} < 0.0$ and $R_{\text{on},i} > 2.0$
- **Transient**: $0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$, $\delta^*_i < 0.4 \mathbb{E}[T_M]$, and $R_{\text{on},i} > 2.0$
- **Sustained**: $0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$, $\delta^*_i > 0.5 \mathbb{E}[T_M]$, and $R_{\text{on},i} > 2.0$

where $\mathbb{E}[T_M] = \alpha_T/\beta_T$ for the Gamma prior, so these thresholds adapt to the chosen hyperparameters.

@@

- **$t^*_{on,i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$**: Absolute activation onset time, typically centered around $0.3 \times \mathbb{E}[T_M]$ to expect activation at 30% of process duration. **Negative values** enable pre-activation scenarios where genes were already activated before observation began.

- **$\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$**: Absolute activation duration, typically centered around $0.4 \times \mathbb{E}[T_M]$ to expect 40% duration activation. This enables both transient (short duration) and sustained (long duration) activation patterns.

@@

- **Pre-activation** ($t^*_{on,i} < 0$): Probability depends on $\mu_{on}$ and $\sigma_{on}$ choices
- **Transient** ($0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$ and $\delta^*_i < 0.4 \mathbb{E}[T_M]$): Balanced probability for observable complete cycles
- **Sustained** ($0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$ and $\delta^*_i > 0.5 \mathbb{E}[T_M]$): Balanced probability for long-duration activation
